% !TeX spellcheck = en_GB

\section*{Acknowledgment}

First of all, special thanks to Dr. Wolfgang Merkle who was willing to be the adviser from my home university, the University of Heidelberg. He introduced me to Prof. Frederik Armknecht from the University of Mannheim and thus laid the foundation for all of this. If not for Merkle, I don't know who else could have been the adviser for a bachelor's thesis in my preferred field, the field of cryptography. Most likely, I would have written my thesis in a different field.

\medskip
\noindent
I am also very grateful to Prof. Frederik Armknecht that he accepted me whom he did not know at all beforehand. He offered me a wide variety of interesting subjects from which I could chose. In the end, I have chosen to develop a plug-in for \textit{CrypTool 2} (CT2), an open-source e-learning platform for cryptography and cryptanalysis.

\medskip
\noindent
Therefore, I want to extend my gratitude to the team behind CT2 whose support and welcomeness meant a lot to me. Prof. Bernhard Esslinger is the overall coordinator and Dr. Nils Kopal is the technical lead, both from the University of Siegen.
Whereas Prof. Armknecht gave me important feedback from a user's perspective since he uses CT2 in his lectures, Prof. Esslinger and Dr. Kopal always took their time to answer technical questions of mine. Additionally, Prof. Esslinger was always eager to remind me about things that were easy to miss like adding text to tooltips which I didn't even know they existed. 

\medskip
\noindent
Finally, I also want to thanks the people at my new employer abusix GmbH. They made it possible for me to focus on my thesis by offering very flexible working hours. I am also very grateful that I could work on my thesis on their premises since going into the office was always a guarantee for a very productive day.
